In the era of big data, efficient data analysis is crucial for extracting valuable insights. On the other hand, the diversity of data systems, file formats, and different analysis types including machine learning models introduce challenges for the quick consolidation of the business data in one data model. To address this issue this paper introduces a cutting-edge concept named Data Virtual Machine (DVM) that revolutionizes data analysis with its user-friendly front-end, flexible data modeling, competitive performance, and innovative data virtualization capabilities. Data Virtual Machine (DVM) is a term used to describe an innovative graph-based conceptual model, resembling an entity-relationship model, that represents relationships between different types of data found in organizational and business settings. This model provides individuals, who lack a technical background but are interested in analyzing data for business-related objectives, the ability to define queries leveraging the adaptable schema and the rapid results enabled by the DVM's capabilities to execute calculations on dataframes seamlessly. Traditionally, data analysis tasks are carried out by experienced data analysts that use procedural approaches which have inherent limitations. With DVMs, the data analysis can be performed visually through an intuitive tool, allowing non-IT experts to participate. Various logical data models can be implemented by the DVM and the focus can be dynamically adjusted to different nodes enabling the analysis of data relationships from different perspectives. Furthermore, DVMs implement data virtualization with the support of data sharing, and portability, and provide a unified view of entities, as each DVM node represents both an attribute and an entity. In this study, we present the optimized version of DataMingler that not only offers ease in data analysis but also delivers rapid results in advanced data analysis cases. This optimization is made possible by leveraging concurrency and parallelization in the calculations performed by DataMingler. Through a comprehensive benchmarking study, we demonstrate the tool's remarkable performance compared to traditional Python implementations.