\documentclass[journal]{IEEEtran} % use the `journal` option for ITherm conference style
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}
\graphicspath{ {./figures/} }
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    urlcolor=black,
    citecolor=black
    }

\urlstyle{same}
%\usepackage{draftwatermark}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}


%\SetWatermarkText{DRAFT} % Watermark text
%\SetWatermarkAngle{45}   % Angle at which watermark appears
%\SetWatermarkLightness{0.93} % Opacity (0 = fully transparent, 1 = fully opaque)
%\SetWatermarkScale{1}    % Watermark scale




\begin{document}

\title{Dataframes Unleashed: Unlocking Effortless Modeling and Performance in Data Analysis with Data Virtual Machines (DVMs)
% delete or comment-out the following line before submission
%{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and should not be used}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{%%%% author names
    \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}% first author
    , \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}% delete this line if not needed
    , \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}% delete this line if not needed
    % duplicate the line above as many times as needed to list all authors
    \\%%%% author affiliations
    \IEEEauthorblockA{\textit{dept. name of organization (of Aff.), City, Country}}\\% first affiliation
    \IEEEauthorblockA{\textit{dept. name of organization (of Aff.), City, Country if needed}}\\% delete this line if not needed
    % duplicate the line above as many times as needed to list all affiliations
    %%%% corresponding author contact details
    \IEEEauthorblockA{email address or ORCID of corresponding author(s)}
}

\maketitle

\begin{abstract}

\input{Sections/ABSTRACT}

\end{abstract}

\begin{IEEEkeywords}
    Dataframes, Data Analysis, Data Virtualization, Graph Data Model, Data Virtual Machines, Efficient Evaluation, User-Friendly Front-End, Big Data, Modeling, Benchmarking, Python, Data Exploration, Insights, Performance
\end{IEEEkeywords}

\section{Introduction}

\input{Sections/1.INTRODUCTION}

\section{Dataframes \& Python}

\input{Sections/2. Dataframes & Python}

\section{Query evaluation \& optimization}

\input{Sections/3.Query evaluation & optimization}

\section{Experiments \& results}

\input{Sections/4.Experiments & results}

\section{Conclusions}

\input{Sections/5.Conclusions}

\newpage
\section{Related Work}

\input{Sections/6.Related Work}

\begin{thebibliography}{00}            
    \bibitem{Aleksic2019}
    Jasna Soldic Aleksic, Biljana Chroneos Krasavac, and Ema Karamata,
    \emph{Business analytics: new concepts and trends},
    \emph{Management: Journal of Sustainable Business and Management Solutions in Emerging Economies}, 
    vol. 25, 
    July 2019,
    doi: \url{10.7595/management.fon.2019.0013}.
    \bibitem{GANDOMI2015137}
    Amir Gandomi and Murtaza Haider,
    \emph{Beyond the hype: Big data concepts, methods, and analytics},
    \emph{International Journal of Information Management}, 
    vol. 35, 
    no. 2, 
    pp. 137-144, 
    2015,
    ISSN 0268-4012,
    doi: \url{https://doi.org/10.1016/j.ijinfomgt.2014.10.007}
    \bibitem{Mauro2016}
    Andrea De Mauro, Marco Greco, Michele Grimaldi, and Giacomo Nobili,
    \emph{Beyond Data Scientists: a Review of Big Data Skills and Job Families},
    in \emph{Proceedings of the Conference Name}, % Replace "Conference Name" with the actual conference name if available
    June 2016.
    \bibitem{Bhadani2016}
    A. Bhadani and D. Jothimani,
    \emph{Big data: Challenges, opportunities and realities},
    In M.K. Singh and D.G. Kumar (Eds.),
    \emph{Effective Big Data Management and Opportunities for Implementation},
    pp. 1-24,
    Pennsylvania, USA,
    IGI Global,
    2016.
    \bibitem{petersohn2020scalable}
    Devin Petersohn, Stephen Macke, Doris Xin, William Ma, Doris Lee, Xiangxi Mo, Joseph E. Gonzalez, Joseph M. Hellerstein, Anthony D. Joseph, and Aditya Parameswaran,
    \emph{Towards Scalable Dataframe Systems},
    \emph{arXiv preprint},
    eprint: 2001.00888,
    archivePrefix: arXiv,
    primaryClass: cs.DB,
    2020.    
    \bibitem{perez2015project}
    Fernando Perez and Brian E. Granger,
    \emph{Project Jupyter: Computational narratives as the engine of collaborative data science},
    Retrieved September, 11(207):108,
    2015.
    \bibitem{abadi2016beckman}
    D. Abadi and others,
    \emph{The Beckman Report on Database Research},
    \emph{Communications of the ACM},
    vol. 59,
    pp. 692–99,
    February 2016.
    \bibitem{chambers1992statistical}
    M. Chambers, T. J. Hastie, and others,
    \emph{Statistical Models in S},
    volume 251,
    Wadsworth \& Brooks/Cole Advanced Books \& Software,
    Pacific Grove, CA,
    1992.
    \bibitem{chambers1990statistical}
    J. Chambers, T. Hastie, and D. Pregibon,
    \emph{Statistical Models in S},
    In K. Momirović and V. Mildner, editors,
    \emph{Compstat},
    pages 317–321,
    Heidelberg,
    1990,
    Physica-Verlag HD.
    \bibitem{McKinney2011}
    Wes McKinney,
    \emph{pandas: a Foundational Python Library for Data Analysis and Statistics},
    \emph{Python High Performance Science Computer},
    January 2011.
    \bibitem{abiteboul1995foundations}
    S. Abiteboul, R. Hull, and V. Vianu,
    \emph{Foundations of Databases},
    volume 8,
    Addison-Wesley,
    Reading,
    1995.
    \bibitem{nelluri2021limitations}
    Ramesh Nelluri,
    \emph{What are the limitations of pandas?},
    \emph{Insights and Data},
    Published on April 10, 2021,
    Available at: \url{https://insightsndata.com/what-are-the-limitations-of-pandas-35d462990c43}.
    \bibitem{sinthong2019aframe}
    Phanwadee Sinthong and Michael J. Carey,
    \emph{AFrame: Extending DataFrames for Large-Scale Modern Data Analysis (Extended Version)},
    \emph{arXiv preprint},
    eprint: 1908.06719,
    archivePrefix: arXiv,
    primaryClass: cs.DB,
    2019.
    \bibitem{chatziantoniou}
    D. Chatziantoniou, V. Kantere, N. Antoniou, A. Gatzia,
    \emph{Data Virtual Machines: Simplifying Data Sharing,
    Exploration \& Querying in Big Data Environments},
    IEEE International Conference on Big Data (Big Data),
    2022.
    
\end{thebibliography}




\end{document}
